---
title: "High dimensional data analysis"
output:
  html_document:
    df_print: paged
---

```{r}
library(ggplot2)
library(heatmaply)
library(Rtsne)
```

```{r}
data = read.csv('geneExpression_GDSC.csv', row.names = "X")
data

summary(data$GeneralType)
summary(data$Type)
```

## Visualize your data with ggplot

You can find a cheat sheet ;).
https://rstudio.com/resources/cheatsheets/

Let's create a bar chat showing the number of cell lines per tissue type.

```{r, fig.width=8, fig.height=8}

ggplot(data, aes(x=GeneralType, fill=GeneralType)) + geom_bar() +
  theme(axis.text.x = element_text(angle=90, hjust=1,vjust=1))

# what happens if you remove theme?

```

Let's try another plot that takes two variables.

```{r, fig.width=8, fig.height=8}

ggplot(data, aes(x=GeneralType, y=JUN, fill=GeneralType)) + geom_boxplot() + #geom_point() +
  theme(axis.text.x = element_text(angle=90, hjust=1,vjust=1))

# you can also try other gene?

```

## Clustering analysis

```{r}
set.seed(1000) # to make sampling reproducible

filter_cell <- data$GeneralType %in% c('breast', 'lung')
filter_gene <- c('Type', 'GeneralType', sample(colnames(data),50))
heatmaply(data[filter_cell, filter_gene], column_text_angle = 90,
          hclust_method = "average") %>% layout(width=500, height=500)
```


## dimensionality reduction (PCA)

```{r, fig.width=7, fig.height=7}

filter_cell <- data$GeneralType %in% c('breast', 'lung', 'pancreas')
PCA = prcomp(data[filter_cell, 3:ncol(data)]) # selecting all numeric columns

barplot(PCA$sdev[1:10]^2, las=2) # amount of variance captured by first 10 components


PCA$x[1:10,1:5]
```

```{r, fig.width=7, fig.height=7}
df <- as.data.frame(PCA$x)
df$Type = data$GeneralType[filter_cell]

ggplot(df, aes(x=PC1, y=PC2, col=Type)) + geom_point()
```

### clustering with the compressed data

```{r }
heatmaply(df[,c("PC1", "PC2", "Type")], column_text_angle = 90,
          hclust_method = "average") %>% layout(width=500, height=500)
```

## Another dimensionality reduction : t-sne

Unlike PCA, t-sne takes local associations than global associations.
(only the closest neighbors considered)
- perplexity parameter controls the number of closest neighbor considered.

Also, t-sne is not reproducible while PCA is.
However, the algorithm gives visually pleasing outcome.

```{r, fig.width=7, fig.height=7}
filter_cell <- data$GeneralType %in% c('breast', 'lung', 'pancreas')
tsne = Rtsne(data[filter_cell, 3:ncol(data)], dims=2, perplexity = 30, max_iter=5000
            ) # selecting all numeric columns

df <- as.data.frame(tsne$Y)
df$Type = data$GeneralType[filter_cell]

ggplot(df, aes(x=V1, y=V2, col=Type)) + geom_point()
```

### clustering with the compressed data

```{r }
heatmaply(df, column_text_angle = 90,
          hclust_method = "average") %>% layout(width=500, height=500)
```